---
title: 'Cyclistic: Members vs Casual Users'
author: "Kevin Camargo"
date: "8/6/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

The goal for this analysis is to find out how annual members and casual riders use Cyclistic bikes
differently.

This analysis will help the marketing team develop a strategy to convert casual riders into annual members.

## Step by Step Analysis Process

### Installing and Loading Packages

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
install.packages('tidyverse', repos = 'http://cran.us.r-project.org')
install.packages('janitor', repos = 'http://cran.us.r-project.org')
install.packages('skimr', repos = 'http://cran.us.r-project.org')
install.packages('formatR', repos = 'http://cran.us.r-project.org')

library(tidyverse)
library(janitor)
library(skimr)
library(readr)
library(ggplot2) 
library(lubridate)
library(formatR)
options(scipen=999) 
```


### Importing Datasets

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
q2_2019 <- read.csv('Cyclistic_Trips_2019_Q2.csv')
q3_2019 <- read.csv('Cyclistic_Trips_2019_Q3.csv')
q4_2019 <- read.csv('Cyclistic_Trips_2019_Q4.csv')
q1_2020 <- read.csv('Cyclistic_Trips_2020_Q1.csv')
```

### Comparing Column Names 

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

### Renaming Columns for Consistency

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
q4_2019 <- rename(q4_2019, ride_id = trip_id, rideable_type = bikeid, started_at = start_time, ended_at = end_time, start_station_name = from_station_name,
                  start_station_id = from_station_id, end_station_name = to_station_name, end_station_id = to_station_id, member_casual = usertype)

q3_2019 <- rename(q3_2019, ride_id = trip_id, rideable_type = bikeid, started_at = start_time, ended_at = end_time, start_station_name = from_station_name,
                  start_station_id = from_station_id, end_station_name = to_station_name, end_station_id = to_station_id, member_casual = usertype)

q2_2019 <- rename(q2_2019, ride_id = X01...Rental.Details.Rental.ID, rideable_type = X01...Rental.Details.Bike.ID, started_at =
                  X01...Rental.Details.Local.Start.Time, ended_at = X01...Rental.Details.Local.End.Time, start_station_name =
                  X03...Rental.Start.Station.Name, start_station_id = X03...Rental.Start.Station.ID, end_station_name = X02...Rental.End.Station.Name,
                  end_station_id = X02...Rental.End.Station.ID, member_casual = User.Type)
```

### Inspecting Dataframes

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```

### Converting ride_id, rideable_type columns to Characters

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
q4_2019 <- mutate(q4_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
q2_2019 <- mutate(q2_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
```

### Stacking All Quarter's Data Into One Dataframe

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

### Removing lat, long, birthyear, and gender Fields 

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, X01...Rental.Details.Duration.In.Seconds.Uncapped
            ,X05...Member.Details.Member.Birthday.Year, Member.Gender, tripduration))
```

### member_casual Column Consolidated From 4 to 2 Labels

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips <- all_trips %>% 
  mutate(member_casual = recode(member_casual, 'Subscriber' = 'member', 'Customer' = 'casual'))
```

### Adding month, day, year, and day_of_week Columns

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), '%m')
all_trips$day <- format(as.Date(all_trips$date), '%d')
all_trips$year <- format(as.Date(all_trips$date), '%Y')
all_trips$day_of_week <- format(as.Date(all_trips$date), '%A')
```

### Adding ride_length Column Calculating ended_at - started_at Columns

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

### Removing ride_length <0 and start_station_name = "HQ QR" (Bad Data)

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == 'HQ QR'| all_trips$ride_length <0),]
```

### ride_length Average, Midpoint, Min, and Max Analysis

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
mean(all_trips_v2$ride_length)  
median(all_trips_v2$ride_length)  
max(all_trips_v2$ride_length)  
min(all_trips_v2$ride_length)
```

### Comparing Average, Midpoint, Min, and Max: Members vs Casual 

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)  
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)  
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)  
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)  
```

###  Average Ride Length by Day: Members vs Casual

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

### Fixing Day of Week Order

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday'
                                                                         ,'Friday', 'Saturday'))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

### Analyzing Data by Type and Weekday

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)
```

### Bar Chart: Number of Rides, Rider Type, Day of Week

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
mindate <- min(all_trips_v2$year)
maxdate <- max(all_trips_v2$year)

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + geom_col(position = 'dodge') + 
  labs(y = 'Number of Rides', x = 'Day of Week', title = 'Members vs Casual Users', subtitle = 'Number of Rides Per Day', fill = 'User Type'
       ,caption = paste0('Data From: ', 'Q2 ', mindate, ' to ', 'Q1 ', maxdate)) + facet_wrap(~member_casual)
```

### Bar Chart: Average Duration, Rider Type, Day of Week 

```{r error=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) + geom_col(position = 'dodge') +
  labs(y = 'Average Ride Duration (sec)', x = 'Day of Week', title = 'Members vs Casual Users', subtitle = 'Average Ride Duration Per Day'
       ,fill = 'User Type', caption = paste0('Data From: ', 'Q2 ', mindate, ' to ', 'Q1 ', maxdate)) + facet_wrap(~member_casual)
```

